Flat-histogram method comparison on the 2D Ising model

### 2020-05-06 ising/ising-body.tex
made a number of ising edits and a thesis edit.

Pommerenck \emph{et al.} propose a refinement~\cite{pommerenck2020stochastic} to SAMC
where the update factor is determined dynamically rather than by the user.
stochastic approximation with a dynamic $\gamma$ (SAD) requires the user to
provide the lowest temperature of interest $T_{\min}$. This is analogous to WL
requiring \emph{a priori.} an energy range of interest; however, this is almost always easier to
identify and is more physical than the SAMC parameter $t_0$. The update factor
can be written in terms of the current estimates for the highest $E_H$ and
lowest $E_L$ energies of interest and the last time that an energy in the range
of interest is encountered $t_L$.
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
SAD only explores the energy range of interest as specified by the minimum
temperature of interest $T_{\min} < T < \infty$. During the simulation the two
energies $E_H$ and $E_L$, are refined such that the range of energies are conservatively
estimated. The weights are calculated for each energy region according to the original
prescription.
\begin{enumerate}
\item {$E < E_L$:} $w(E>E_H) = w(E_H)$
\item {$E_L < E < E_H$:} moves are handled the same as other weight-based
methods that are mentioned
\item {$E > E_H$:} $w(E<E_L) = w(E_L)e^{-\frac{E_L-E}{T_{\min}}}$
\end{enumerate}
Each time the simulation changes the value of $E_H$ or $E_L$, the weights
within the new portion of the interesting energy range are updated.

\section{Results}
We tested the algorithms on two different system sizes of the 2D Ising model.  The first is a smaller
simulation with a lattice size of $N = 32 \times 32$ and the second has a lattice size of $N = 128 \times 128$. The SAD method explores the energy
space of each system using a minimum reduced temperature of $T_{\text{min}} = 1$. All simulations lead to the minimum important energy $E_{\min}$
and maximum entropy energy $E_{\max}$ being calculated (with the exception of
the WL methods where both of these parameters are needed \emph{a priori.}).

@@

All the methods, except for SAMC, use the same single random number seed and
give a reasonable approximation for the heat capacity peak resulting from the
phase transition within 10$^9$ moves. Figure~\ref{fig:N32-cv} shows the
specific heat capacity vs. the reciprocal temperature $\beta$ at $10^{9}$ moves.
A temperature range of $1.5k_B$ to $5k_B$ is chosen to highlight
the phase transition at the critical temperature $T_c$.

@@

\subsection{128 $\times$ 128 Ising system}
For the larger 2D Ising system, the update factors initially
start at the same $\gamma_{t=0} = 1$; however, all of the methods take longer
to proceed as $1/t$ (with SAD taking $10^{13}$ moves). Fig.~\ref{fig:N128-gamma} shows the update factor $\gamma_t$ for each of the 
flat-histogram methods. 
We implement two WL simulations followed by a production run each beginning
after $\gamma^{\text{WL}}_t$ has reached $10^{-4}$ and $10^{-6}$ respectively.
Fig.~\ref{fig:N128-cv-error} shows the maximum error in the heat capacity as a 
function of time for this system. The solid/dashed lines represent the average
of the maximum value of the error in the specific heat capacity $c_V$ averaged
over eight simulations using different random number seeds. The range of maximum
errors for each simulation is shown as a shaded region. By
the time $10^{13}$ moves have been made all but the WL simulation have begun to
converge as $1/\sqrt{t}$. We then see the WL error saturate around $10^{12}$
moves. The two WL simulations followed by a production run highlight that
choosing too large a $\gamma^{\text{WL}}_t$ will result in poor convergence
properties.

\section{Conclusion}
We find that both SAD, 1/t-WL, and WL followed by a production run
(with an adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure WL do
not suffer from
error saturation. We find that for larger systems SAD reduces the updated factor
more slowly (and conservatively) than 1/t-WL and WL followed by a
production run. This means that SAD will take
proportionately more moves to converge to the same value as 1/t-WL as system
size is increased. SAD requires the user to specify a temperature range of
interest rather than an energy range of interest as the WL methods do. The 2D
Ising model is a system where the energy range is known \emph{a priori.}, which
makes a comparison between WL methods and SAD (which requires a minimum
temperature) more biased towards WL methods. For the majority of cases in which
the energy range of the system is not known and where a range of desired
temperatures is known, this will make the SAD method
considerably more convenient. This work also visually demonstrates
that WL followed by a production run performs extremely well and is preferable
to WL for ensuring both ergodicity and detailed balance.

>>>>

Pommerenck \emph{et al.} propose a refinement~\cite{pommerenck2020stochastic} to SAMC
where the update factor is determined dynamically rather than by the user.
Stochastic approximation with a dynamic $\gamma$ (SAD) requires the user to
provide a minimum temperature of interest $T_{\min}$. This is analogous to WL
requiring \emph{a priori.} an energy range of interest; however, this is almost always easier to
identify and is more physical than the SAMC parameter $t_0$. The update factor
can be written in terms of the current estimates for the highest $E_H$ and
lowest $E_L$ interesting energies and the last time that an energy in the range
of interest is encountered $t_L$.
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
SAD only explores the energy range of interest as specified by the minimum
temperature of interest $T_{\min} < T < \infty$. During the simulation the two
energies $E_H$ and $E_L$, are refined such that the range of energies are
conservatively estimated. The weights are calculated for each energy region according to the original prescription.
\begin{enumerate}
\item {$E < E_L$:} $w(E>E_H) = w(E_H)$
\item {$E_L < E < E_H$:} moves are handled the same as other weight-based
methods that are mentioned
\item {$E > E_H$:} $w(E<E_L) = w(E_L)e^{-\frac{E_L-E}{T_{\min}}}$
\end{enumerate}
Each time the simulation changes the value of $E_H$ or $E_L$, the weights
within the new portion of the interesting energy range are updated.

\section{Results}
We test the algorithms on two different system sizes of the 2D Ising model.
The first is a smaller simulation with a lattice size of $N = 32 \times 32$ and
the second has a lattice size of $N = 128 \times 128$. The SAD method explores 
the energy space of each system using a minimum reduced temperature of $T_{\text{min}} = 1$. All simulations calculate the minimum important energy $E_{\min}$
and maximum entropy energy $E_{\max}$ (with the exception of the WL methods
where both of these parameters are needed \emph{a priori.}).

@@

All of the methods, except for SAMC, use the same single random number seed and
give a reasonable approximation for the heat capacity peak resulting from the
phase transition after only 10$^9$ moves. Figure~\ref{fig:N32-cv} shows the
specific heat capacity vs. the reciprocal temperature $\beta$ at $10^{9}$ moves.
A temperature range of $1.5k_B$ to $5k_B$ is chosen to highlight
the phase transition at the critical temperature $T_c$.

@@

\subsection{128 $\times$ 128 Ising system}
For the larger 2D Ising system, the update factors initially
start at the same $\gamma_{t=0} = 1$; however, all of the methods take longer
to proceed as $1/t$ (with SAD taking $10^{13}$ moves). Fig.~\ref{fig:N128-gamma} shows the update factor $\gamma_t$ for each of the 
flat-histogram methods. 
We implement two WL simulations followed by a production run each beginning
after $\gamma^{\text{WL}}_t$ has reached $10^{-4}$ and $10^{-6}$ respectively.
Fig.~\ref{fig:N128-cv-error} shows the maximum error in the heat capacity as a 
function of time for this system. The solid/dashed lines represent the maximum
value of the error in the specific heat capacity $c_V$ averaged
over eight simulations using different random number seeds. The range of maximum
errors for each simulation is shown as a shaded region. By
the time $10^{13}$ moves have been made all but the WL simulation have begun to
converge as $1/\sqrt{t}$. We then see the WL error saturate around $10^{12}$
moves. The two WL simulations followed by a production run highlight that
choosing too large a $\gamma^{\text{WL}}_t$ will result in poor convergence
properties.

\section{Conclusion}
We find that both SAD, 1/t-WL, and WL followed by a production run
(with an adequately small $\gamma^{\min}$) demonstrate excellent and robust
convergence. They all converge more rapidly than SAMC, and unlike pure WL do
not suffer from
error saturation. We find that for larger systems SAD reduces the updated factor
more slowly (and conservatively) than 1/t-WL and WL followed by a
production run. This means that SAD will take
proportionately more moves to converge to the same value as 1/t-WL as system
size is increased. SAD requires the user to specify a temperature range of
interest rather than an energy range of interest as the WL methods do. The 2D
Ising model is a system where the energy range is known \emph{a priori.}, which
makes a comparison between WL methods and SAD (which requires a minimum
temperature) more biased towards WL methods. For the majority of cases in which
the energy range of the system is not known and where a range of desired
temperatures is known, this will make the SAD method
considerably more convenient. This work also visually demonstrates
that WL followed by a production run performs extremely well and is preferable
to pure WL for ensuring both ergodicity and detailed balance.
